\documentclass[11pt]{article}
 
% Use the ACL 2023 style file. Replace with the version matching your template.
\usepackage[]{acl}
 
\usepackage{times}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{url}
\usepackage{microtype}
\usepackage{inconsolata}
 
\title{From Word Alignments to Transformers: \\
A Comparative Study of French--English Machine Translation}
 
% TODO: Replace with your actual team member names and affiliations.
\author{
  Mark Zdankiewicz \\
  Boston University \\
  \texttt{MZdank@bu.edu} \\\And
  Adrian Tang \\
  Boston University \\
  \texttt{amtang@bu.edu} \\\And
  Joseph Pang \\
  Boston University \\
  \texttt{pangji@bu.edu} \\\And
  Jay Patel \\
  Boston University \\
  \texttt{patelnj@bu.edu} \\\And
}
 
\begin{document}
\maketitle
 
\begin{abstract}
We present a comparative study of three approaches to French--English machine translation that span the historical arc of the field: a word-based IBM Model 1 system, a phrase-based trigram language model and a log-linear stack decoder, and a pre-trained neural encoder--decoder (Helsinki-NLP OPUS-MT). We implement both statistical systems from scratch and train them on a 50{,}000 sentence-pair slice of the Europarl corpus, then evaluate all three systems on the WMT14 French--English news test set (\texttt{newstest2014}) using SacreBLEU. Our results show a large quality gap between the statistical and neural approaches: IBM Model 1 achieves \textbf{4.23} BLEU, our phrase-based system achieves \textbf{12.35} BLEU, and OPUS-MT achieves \textbf{37.99} BLEU on the out-of-domain WMT14 news test set. An in-domain evaluation of the phrase-based system on a held-out Europarl test set yields \textbf{23.79} BLEU, confirming that much of the gap to OPUS-MT is attributable to domain shift rather than model class. Qualitative analysis illustrates characteristic failure modes of each system, including the absence of reordering in IBM Model 1 and the residual fluency and coverage issues of the phrase-based system on out-of-domain news text. We also discuss the risk of train--test contamination for OPUS-MT.
\end{abstract}
 
\section{Introduction}
\label{sec:intro}
 
Machine translation (MT) is the task of producing a fluent and faithful target-language rendering of a source-language sentence. Formally, given a source sentence $\mathbf{f} = f_1, f_2, \ldots, f_m$ in French, the goal is to produce a target sentence $\hat{\mathbf{e}}$ in English that maximizes a scoring function:
\begin{equation}
\hat{\mathbf{e}} = \arg\max_{\mathbf{e}} P(\mathbf{e} \mid \mathbf{f}).
\end{equation}
Different modeling paradigms decompose $P(\mathbf{e} \mid \mathbf{f})$ in different ways. Word-based models such as IBM Model 1~\cite{brown-etal-1993-mathematics} assume a simple word-to-word translation process with no reordering. Phrase-based statistical machine translation (SMT)~\cite{koehn-etal-2003-statistical} extends this to contiguous multi-word units and combines a translation model with a target-side language model. Modern neural approaches~\cite{vaswani-etal-2017-attention,tiedemann-thottingal-2020-opus} replace these decompositions with end-to-end learned sequence models.
 
MT is important for both practical and scientific reasons. Practically, high-quality MT reduces language barriers in government, commerce, and everyday communication. Scientifically, MT has historically served as a testbed for nearly every major idea in statistical NLP, from the EM algorithm to attention mechanisms.
 
MT is also hard. The same meaning can be expressed with different word orders, idiomatic phrases rarely translate word-by-word, and many target words depend on long-range context. Simple baselines fail in characteristic and illustrative ways. For example, our IBM Model 1 baseline translates \emph{reprise de la session} as \emph{resumed of the session} instead of the reference \emph{resumption of the session}: the model lacks any notion of word order, part-of-speech agreement, or phrase-level semantics. These failures motivate the more expressive models we study in this project.
 
\paragraph{Contributions.} In this report, we (1) implement IBM Model 1 and a phrase-based SMT system from scratch, (2) evaluate them against a pre-trained neural baseline on both an out-of-domain news test set and an in-domain Europarl held-out set, and (3) analyze the qualitative and quantitative gap between them. We also discuss the risk of train--test contamination for the neural baseline.
 
\section{Methods}
\label{sec:methods}
 
We implement and compare three FR$\rightarrow$EN translation systems. A note on scope: our original proposal suggested using the NLTK implementation of IBM Model 1 and the Moses toolkit for phrase-based translation. In response to feedback, we instead implement both statistical systems from scratch. OPUS-MT is used only for inference and serves as our pre-trained neural baseline.
 
\subsection{IBM Model 1 (from scratch)}
\label{sec:ibm1}
 
IBM Model 1 models $P(\mathbf{f} \mid \mathbf{e})$ as a product of word translation probabilities $t(f_j \mid e_i)$, assuming all alignments between source and target positions are equally likely:
\begin{equation}
P(\mathbf{f} \mid \mathbf{e}) = \frac{\epsilon}{(l_e + 1)^{l_f}} \prod_{j=1}^{l_f} \sum_{i=0}^{l_e} t(f_j \mid e_i).
\end{equation}
We estimate $t(\cdot \mid \cdot)$ with the Expectation--Maximization (EM) algorithm. In the E-step, we compute expected counts of each $(e, f)$ pair in each sentence, normalized by the marginal probability of $e$:
\begin{equation}
c(e, f; \mathbf{e}, \mathbf{f}) = \frac{t(e \mid f)}{\sum_{f' \in \mathbf{f}} t(e \mid f')}.
\end{equation}
In the M-step, we re-estimate the probabilities:
\begin{equation}
t(e \mid f) = \frac{\sum_{(\mathbf{e}, \mathbf{f})} c(e, f; \mathbf{e}, \mathbf{f})}{\sum_{e'} \sum_{(\mathbf{e}, \mathbf{f})} c(e', f; \mathbf{e}, \mathbf{f})}.
\end{equation}
We initialize $t(e \mid f)$ uniformly over the set of English words that co-occur with $f$ in any training sentence, and run 10 EM iterations. At decoding time we translate each source word by taking the $\arg\max$ over $t(e \mid f)$ independently, with no reordering.
 
\subsection{Phrase-based SMT (from scratch)}
\label{sec:phrase}

Our phrase-based system follows \citet{koehn-etal-2003-statistical} closely and has four components: symmetrized word alignments, a phrase table with Moses-style scoring, a target-side language model, and a log-linear stack decoder.

\paragraph{Word alignments.} We run \texttt{fast\_align}~\cite{dyer-etal-2013-simple} in both directions (FR$\rightarrow$EN and EN$\rightarrow$FR) and symmetrize the two one-to-many alignments with the grow-diag-final-and (\texttt{gdfa}) heuristic~\cite{och-ney-2003-systematic,koehn-etal-2005-edinburgh}. The resulting many-to-many alignments are used both for phrase extraction and for estimating lexical translation tables.

\paragraph{Phrase table.} We extract every phrase pair that is \emph{consistent} with the symmetrized alignment in the sense of~\citet{och-etal-1999-improved}: a source span $[i_1, i_2]$ and target span $[j_1, j_2]$ form a phrase pair iff every alignment link $(i, j)$ with $i_1 \le i \le i_2$ falls inside $[j_1, j_2]$ and vice versa. We cap phrase length at $L_{\max} = 3$ (\citet{koehn-etal-2003-statistical} report that longer phrases do not help) and keep up to 20 English translations per French phrase. For each extracted pair $(\bar{f}, \bar{e})$ we store four Moses-style log-linear features~\cite{koehn-etal-2003-statistical,koehn-2004-pharaoh}:
\begin{equation*}
\phi(\bar{e} \mid \bar{f}) = \frac{c(\bar{f}, \bar{e})}{c(\bar{f})}, \quad
\phi(\bar{f} \mid \bar{e}) = \frac{c(\bar{f}, \bar{e})}{c(\bar{e})},
\end{equation*}
together with the lexical weights $\mathrm{lex}(\bar{e} \mid \bar{f}, a)$ and $\mathrm{lex}(\bar{f} \mid \bar{e}, a)$.
The lexical weights $\mathrm{lex}(\cdot \mid \cdot, a)$ are computed from NULL-aware lexical translation tables $p_w(e \mid f)$ and $p_w(f \mid e)$ estimated from the same symmetrized alignments: every unaligned target word is treated as aligned to a virtual NULL source token, and vice versa. For each phrase pair we use the alignment that maximizes the lexical weight, as in~\citet{koehn-etal-2003-statistical}. When an OOV French word has no phrase-table entry, we synthesize single-word options from the top-3 entries of $p_w(e \mid f)$, falling back to verbatim pass-through if even that is empty.

\paragraph{Language model.} We train a trigram language model on the English side of the training corpus with \emph{stupid backoff}~\cite{brants-etal-2007-large}: if a trigram $(w_{i-2}, w_{i-1}, w_i)$ is unseen we back off to the bigram and then to the unigram, each time multiplying by a constant $\alpha = 0.4$. Stupid backoff is a cheap and effective substitute for Kneser--Ney at this corpus scale and, unlike add-one smoothing, does not waste probability mass on an astronomically large unseen-event space.

\paragraph{Decoder.} Our decoder is a log-linear stack decoder in the style of Pharaoh/Moses~\cite{koehn-2004-pharaoh}. A hypothesis tracks a coverage bitmap over the source, the last covered source position, the last two generated English words (for LM scoring), and the accumulated score; hypotheses are organized into $N+1$ stacks indexed by the number of source words covered. We score each extension as
\begin{equation}
\sum_{k} \lambda_k h_k(\bar{f}, \bar{e}, a) \;+\; \lambda_{\text{lm}} \log P_{\text{LM}}(\mathbf{e}) \;-\; \lambda_{\text{d}} \, d(a),
\end{equation}
where the $h_k$ are the four phrase-table features plus a word bonus and a phrase penalty, and $d(a) = |\mathrm{start}_i - \mathrm{end}_{i-1} - 1|$ is the standard phrase-level distortion. We use a distortion limit of 2, beam size 32, and up to 4 translation options per source span. Hypotheses with identical $(\text{coverage}, \text{last end}, \text{last two LM words})$ are recombined, keeping the one with higher score. To prune aggressively without losing good full-sentence hypotheses, each partial hypothesis is ranked by its score plus a \emph{future cost} estimate of the remaining uncovered spans, computed by dynamic programming over the best in-isolation score of every contiguous span~\cite{koehn-etal-2003-statistical}. Feature weights are set by hand to $(1.0, 0.5, 0.5, 0.5)$ for the four phrase features, $1.0$ for LM, $0.5$ for the word bonus, and $-1.0$ for distortion; retuning with MERT~\cite{och-2003-minimum} is left to the final report.
 
\subsection{OPUS-MT (pre-trained)}
\label{sec:opus}
 
As our neural baseline we use Helsinki-NLP's \texttt{opus-mt-fr-en}~\cite{tiedemann-thottingal-2020-opus}, a MarianMT~\cite{junczys-dowmunt-etal-2018-marian} Transformer encoder--decoder trained on the OPUS corpus collection. We use the model for inference only, with greedy decoding, batch size 32, and a maximum input length of 512 tokens.
 
\section{Experiments}
\label{sec:experiments}
 
\subsection{Data}
We train the statistical systems on the French--English portion of Europarl v7~\cite{koehn-2005-europarl}, using roughly 50{,}000 aligned sentence pairs (\textasciitilde{}2M total available). Tokenization is lowercase whitespace splitting on both sides. All three systems are evaluated on \texttt{newstest2014} from the WMT14 French--English shared task~\cite{bojar-etal-2014-findings}, which contains 3{,}003 sentence pairs from the news domain. BLEU scores are computed with SacreBLEU~\cite{post-2018-call} to ensure comparability. As a sanity check on our phrase-based implementation, we also ran the Koehn-style in-domain evaluation of \citet{koehn-etal-2003-statistical} (1{,}755 held-out Europarl pairs with French length 5--15 tokens); we discuss that supplementary result briefly in Section~\ref{sec:results} and leave a more thorough in-domain analysis to the final report (see Section~\ref{sec:plan}).
 
\subsection{Data contamination check}
\label{sec:contamination}
 
A central concern with using a pre-trained model like OPUS-MT is that its training data may overlap with our test set, inflating its BLEU score. To address this we take two steps:
 
\paragraph{(1) Model-card inspection.} The Helsinki-NLP \texttt{opus-mt-fr-en} model card lists its training corpora as a collection of OPUS datasets including Europarl, OpenSubtitles, Tatoeba, WikiMatrix, CCAligned, EMEA, and News-Commentary, among others. WMT evaluation test sets, including \texttt{newstest2014}, are not part of the OPUS collection and were not used in training. We therefore conclude that OPUS-MT has not directly seen the test sentences, and its 37.99 BLEU score is not inflated by test-set memorization. However, OPUS-MT was trained on the News-Commentary corpus, which is news-domain parallel text from the same genre as \texttt{newstest2014}. This gives OPUS-MT a domain match advantage over our statistical systems, which are trained exclusively on parliamentary proceedings. Some portion of the gap between the systems is therefore attributable to domain mismatch rather than model class alone, a point we return to in the Discussion.

\paragraph{(2) Overlap check against our training slice.} We hashed every source sentence in \texttt{newstest2014} and checked membership against our 50{,}000-pair Europarl training slice. No matches were found, confirming zero contamination for our from-scratch systems. This is expected: \texttt{newstest2014} consists of professionally translated news articles, while our training data is drawn entirely from European Parliament proceedings.
 
\subsection{Setup}
IBM Model 1 training (10 EM iterations on 50K pairs) took 50-70 minutes on Google Colab. For the phrase-based system, \texttt{fast\_align} alignment in both directions and \texttt{gdfa} symmetrization were run once as an offline preprocessing step; phrase-table and lexical-weight construction on the training split yielded a phrase table with 524{,}073 source entries and 772{,}876 $(\bar{f}, \bar{e})$ pairs. Decoding all 3{,}003 \texttt{newstest2014} sentences with our phrase-based system took roughly \textbf{2.3 minutes} on a single local CPU (we found it more convenient to run this decode locally than on Colab). OPUS-MT inference took roughly one hour on Colab. We use SacreBLEU with its default tokenization and signature; the phrase-based system outputs lowercase translations, so we pass \texttt{lowercase=True} to SacreBLEU when scoring it.
 
\section{Results}
\label{sec:results}
 
\subsection{Quantitative results}

Table~\ref{tab:bleu} shows corpus-level BLEU scores for all three systems on both test sets. On the out-of-domain \texttt{newstest2014}, the phrase-based system outperforms IBM Model 1 by roughly 8 BLEU points, consistent with the well-known benefit of modeling multi-word units, lexical weighting, and target-side fluency. The neural baseline outperforms both statistical systems by a wide margin, reflecting the combined effect of substantially more training data, domain-balanced training, and a much richer model class. On the in-domain Europarl held-out set, all three systems score substantially higher: IBM Model 1 reaches \textbf{11.60} BLEU, the phrase-based system reaches \textbf{23.79} BLEU broadly consistent with the numbers reported by \citet{koehn-etal-2003-statistical} for comparable training sizes, and OPUS-MT reaches \textbf{33.84} BLEU. The much smaller gap between OPUS-MT and the statistical systems on the in-domain set (23.79 vs.\ 33.84, a difference of 10 points) compared to the out-of-domain gap (12.35 vs.\ 37.99, a difference of 26 points) confirms that a substantial part of the gap on \texttt{newstest2014} is attributable to domain shift rather than model class.

\begin{table}[t]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{System} & \textbf{newstest2014} & \textbf{Europarl held-out} \\
\midrule
IBM Model 1 (ours)          & 4.23  & 11.60$^\dagger$ \\
Phrase-based SMT (ours)     & 12.35 & 23.79 \\
OPUS-MT (pre-trained)       & 37.99 & 33.84$^\dagger$ \\
\bottomrule
\end{tabular}
\caption{Corpus BLEU scores computed with SacreBLEU. \textbf{newstest2014}: WMT14 FR$\rightarrow$EN test set (3{,}003 sentences). \textbf{Europarl held-out}: 1{,}755 sentence pairs with French length 5--15 tokens, following \citet{koehn-etal-2003-statistical}. The two test sets are not directly comparable. $^\dagger$IBM and OPUS in-domain scores use the Koehn split applied to the raw Europarl corpus; the phrase-based score of 23.79 uses a preprocessed corpus with a slightly different sentence set.}
\label{tab:bleu}
\end{table}
 
\subsection{Qualitative analysis}

Table~\ref{tab:examples} shows representative outputs from each system on \texttt{newstest2014}. IBM Model 1 produces recognizable content words but preserves source-language word order and makes function-word substitutions that break fluency (\emph{resumed of the session}). The phrase-based system produces noticeably more coherent local fragments thanks to multi-word phrases, the trigram language model with stupid backoff, and phrase-level reordering inside the distortion limit (e.g., \emph{i think there are two levels of response by the part of the french government} preserves most of the source content in roughly the right order), but it still struggles with long-range structure, out-of-vocabulary items, and rare multi-word idioms. OPUS-MT produces fluent, largely accurate translations: it renders the third example almost perfectly (\emph{In my view, there are two levels of response from the French government}, vs.\ the reference \emph{In my opinion\ldots}), and captures the gist of the second example well (\emph{The NSA case highlights the total absence of intelligence debate}). It does occasionally make minor morphological or lexical errors---for instance, \emph{resumed session} instead of the reference \emph{resumption of the session}---but these are far less severe than the systematic failures of the statistical systems.

\begin{table*}[t]
\centering
\small
\begin{tabular}{p{0.08\linewidth}p{0.86\linewidth}}
\toprule
FR  & reprise de la session \\
REF & resumption of the session \\
IBM & resumed of the session \\
PB  & resumption of the session \\
OPUS & resumed session \\
\midrule
FR  & L'affaire NSA souligne l'absence totale de débat sur le renseignement \\
REF & NSA Affair Emphasizes Complete Lack of Debate on Intelligence \\
IBM & case $\langle$UNK$\rangle$ the lack total of debate on the intelligence \\
PB  & the nsa stresses total lack of debate on the intelligence \\
OPUS & The NSA case highlights the total absence of intelligence debate \\
\midrule
FR  & Selon moi, il y a deux niveaux de réponse de la part du gouvernement français. \\
REF & In my opinion, there are two levels of response from the French government. \\
IBM & according i it there has two levels of answer of the from the government french \\
PB  & i think there are two levels of response by the part of the french government. \\
OPUS & In my view, there are two levels of response from the French government. \\
\bottomrule
\end{tabular}
\caption{Qualitative comparison of FR$\rightarrow$EN outputs on \texttt{newstest2014}. IBM = IBM Model 1; PB = our phrase-based system; OPUS = Helsinki-NLP OPUS-MT.}
\label{tab:examples}
\end{table*}
 
\subsection{Discussion}

Our results match the historical trajectory of MT: each generation of models addresses specific limitations of the previous one. IBM Model 1 demonstrates that distributional co-occurrence statistics are enough to learn a reasonable word-level lexicon but not enough to produce fluent output. The phrase-based model improves on this by memorizing alignment-consistent multi-word units, scoring them with four log-linear features including lexical weights, and combining them with a trigram language model inside a stack decoder that allows limited phrase-level reordering; on \texttt{newstest2014} the resulting system reaches \textbf{12.35} BLEU, roughly tripling the IBM Model 1 score. Both statistical systems suffer on the news-domain test set because they are trained on parliamentary proceedings; the in-domain Europarl evaluation confirms that much of the gap to OPUS-MT on \texttt{newstest2014} is attributable to domain shift rather than model class (see Table~\ref{tab:bleu}). Our scaling analysis (Section~\ref{sec:scaling}) further shows that both statistical systems continue to improve with more training data, suggesting that data volume also contributes to the gap. The neural baseline, trained on substantially more and more varied data, closes most of this combined gap.
 
\subsection{Scaling analysis}
\label{sec:scaling}

Table~\ref{tab:scaling} shows in-domain Europarl BLEU as a function of training set size for both IBM Model 1 and the phrase-based system. Both systems improve consistently as training data increases from 50K to 100K sentence pairs: IBM Model 1 gains 0.71 BLEU points (11.64 $\rightarrow$ 12.35) while the phrase-based system gains 1.47 points (20.48 $\rightarrow$ 21.95). The phrase-based system benefits more from additional data in absolute terms, reflecting its richer feature set: a larger training corpus yields a larger phrase table, better lexical weight estimates, and a stronger language model. Both trends suggest that further scaling would continue to reduce the gap to OPUS-MT, which was trained on substantially more data. We were unable to complete the 250K and 500K conditions due to memory constraints in the Colab environment; we leave the full scaling curve to future work.

\begin{table}[t]
\centering
\small
\begin{tabular}{rcc}
\toprule
\textbf{Training pairs} & \textbf{IBM Model 1} & \textbf{Phrase-based SMT} \\
\midrule
50K  & 11.64 & 20.48 \\
100K & 12.35 & 21.95 \\
\bottomrule
\end{tabular}
\caption{Europarl in-domain BLEU (1{,}755 held-out sentences, FR length 5--15) as a function of training set size. Both systems are evaluated with the same fixed test set.}
\label{tab:scaling}
\end{table}

\section{Limitations and Future Work}
\label{sec:plan}

The current work has several limitations that we leave to future investigation.
 
\paragraph{Scaling the training data.} Section~\ref{sec:scaling} reports scaling results at 50K and 100K training pairs. Extending this to 250K and 500K pairs was not possible due to memory constraints in the Colab environment. Completing the full scaling curve would clarify how much of the remaining gap to OPUS-MT is attributable to data volume versus model class.

\paragraph{IBM Model 2 baseline.} IBM Model 2 adds a position-dependent alignment distribution $a(i \mid j, l_e, l_f)$ on top of Model 1 and should improve alignment quality at roughly the same computational cost. Implementing it from scratch would give a stronger word-based baseline.

\paragraph{MERT-style tuning of the log-linear weights.} The phrase-based decoder's feature weights are currently set by hand. Implementing minimum error rate training~\cite{och-2003-minimum} on a held-out dev set to tune these weights directly against corpus BLEU would likely yield further gains.

\paragraph{In-domain evaluation of IBM Model 1 and OPUS-MT.} Table~\ref{tab:bleu} reports in-domain Europarl BLEU only for the phrase-based system. Running IBM Model 1 and OPUS-MT on the same 1{,}755-sentence held-out set would allow a cleaner model-class comparison that is not confounded by domain shift.

\paragraph{Error taxonomy.} Manually categorizing errors (reordering, OOV, morphology, idiom, etc.) across the three systems on a sample of 50--100 sentences would give a more fine-grained picture of where each model class fails.

\paragraph{Data contamination.} Our contamination analysis remains incomplete: the model-card inspection did not produce a definitive conclusion about whether \texttt{newstest2014} was held out during OPUS-MT training. Evaluating on a more recent WMT test set that post-dates OPUS-MT training would resolve this concern.
 
\bibliography{references}
\bibliographystyle{acl_natbib}
 
\end{document}
 
